---
title: "LAB 2"
author: "Andreas Hertin"
date: '2021-09-16'
output: word_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
set.seed(10)
library(HMM)

```

The purpose of the lab is to put in practice some of the concepts covered in the lectures.
To do so, you are asked to model the behavior of a robot that walks around a ring. The ring is
divided into 10 sectors. At any given time point, the robot is in one of the sectors and decides
with equal probability to stay in that sector or move to the next sector. You do not have direct
observation of the robot. However, the robot is equipped with a tracking device that you can
access. The device is not very accurate though: If the robot is in the sector i, then the device
will report that the robot is in the sectors [i − 2, i + 2] with equal probability

## (1)

Build a hidden Markov model (HMM) for the scenario described above.

```{r}

states = 1:10
symbols = 1:10

# Create transition matrix to match description
transMatrix <- diag(0.5, 10)
transMatrix[10, 1] = 0.5
diag(transMatrix[, -1]) = 0.5

emissMatrix <- matrix(0,10,10)

startMatrix <- rep(0.1,10)


for(i in 1:10){
  # get Y values that should be filled with 0.2
  lowerBound = i - 2
  upperBound = i + 2
  y_fill = lowerBound:upperBound%%10
  # Change 0 indexes to 10
  y_fill[y_fill == 0] = 10
  # Fill matrix by (i,y_fill) indexes with 0.2
  emissMatrix[i, y_fill] = 0.2
}

transMatrix
emissMatrix
startMatrix

hmmModel <- initHMM(States = states, Symbols = symbols, startProb = startMatrix, transProbs = transMatrix, emissionProbs = emissMatrix)
hmmModel

```


## (2)

Simulate the HMM for 100 time steps.

```{r}

N = 100

hmmModel.sim <- simHMM(hmm = hmmModel, length = N)

hmmModel.sim$states
hmmModel.sim$observation

```


## (3)

Discard the hidden states from the sample obtained above. Use the remaining observations to compute the filtered and smoothed probability distributions for each of the
100 time points. Compute also the most probable path.

```{r}

alpha <- exp(forward(hmmModel, hmmModel.sim$observation))
beta <- exp(backward(hmmModel, hmmModel.sim$observation))

filterFunc = function(HMM, a, N){
  out = matrix(0, 10, N)
  for(i in 1:N){
    out[, i] = a[, i] / sum(a[, i])
  }
  return(out)
}

smoothFunc = function(HMM, a, b, N){
  out = matrix(0, 10, N)
  for(i in 1:N){
    out[, i] = b[, i] * a[, i] / sum(b[, i] * a[, i])
  }
  return(out)
}

filtered <- filterFunc(hmmModel, alpha, N)
smoothed <- smoothFunc(hmmModel.sim, alpha, beta, N)
mostProbablePath <- viterbi(hmmModel, hmmModel.sim$observation)
mostProbablePath

```


## (4)

Compute the accuracy of the filtered and smoothed probability distributions, and of the
most probable path. That is, compute the percentage of the true hidden states that are
guessed by each method.

**Hint**: Note that the function ``forward()`` in the ``HMM`` package returns probabilities in log
scale. You may need to use the functions ``exp()`` and ``prop.table()`` in order to obtain a
normalized probability distribution. You may also want to use the functions ``apply()`` and
``which.max()`` to find out the most probable states. Finally, recall that you can compare
two vectors A and B elementwise as A==B, and that the function ``table()`` will count the
number of times that the different elements in a vector occur in the vector.


```{r}



```




