---
title: "Lab4"
author: "Andreas Hertin"
date: '2021-10-10'
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("mvtnorm")
# install.packages('kernlab')
library("mvtnorm")
library(kernlab)
library(scales)
```


## (1)
Implementing GP Regression

## (1.1)
First task is implementing a function that derives the posterior distribution from given prior points and training data.
```{r}
# Inspired by given code from José
SquaredExpKernel <- function(x1,x2,sigmaF=1,l=3){
  n1 <- length(x1)
  n2 <- length(x2)
  K <- matrix(NA,n1,n2)
  for (i in 1:n2){
    K[,i] <- sigmaF^2*exp(-0.5*( (x1-x2[i])/l)^2 )
  }
  return(K)
}


posteriorGP <- function(x, y, XStar, sigmaNoise, k, sigmaF=1, l=3){
  n <- length(x)
  K <- k(x, x, sigmaF, l)
  L <- t(chol(K + sigmaNoise^2*diag(n)))
  L_t <- t(L)
  alpha <- solve(L_t, solve(L, y))
  K_star <- k(x, XStar, sigmaF, l)
  f_star <- t(K_star) %*% alpha
  
  v <- solve(L, K_star)
  K_star_star <- k(XStar, XStar, sigmaF, l)
  cov <- K_star_star - t(v) %*% v
  var <- diag(cov)
  
  return(list(mean = f_star, var = var))
  
}

```

## (1.2)
In this section the function is run with $\sigma_f = 1$ and $l = 0.3$ and a single  point.
```{r}

plotGP <- function(mean, var = NULL, interval, x, y, title){
  
  if(!is.null(var)){
  
    conf_int <- list(upper = mean + 1.96 * sqrt(var),
                     lower = mean - 1.96 * sqrt(var))
    
    ylim <- c(min(conf_int$lower) - 2,
              max(conf_int$upper) + 2)
  
    plot(interval, 
         mean, 
         type = "l", 
         ylim = ylim, 
         col = "red", 
         main = title, 
         xlab = "x", 
         ylab= "Posterior mean",
         lwd = 2)
    
    lines(x = interval, y = conf_int$lower, col = "black", lwd = 2, lty = 2)
    lines(x = interval, y = conf_int$upper, col = "black", lwd = 2, lty = 2)
    
  } else {
    
    ylim <- c(min(y) - 1,
              max(y) + 1)
    
    plot(interval, 
         mean, 
         type = "l", 
         main = title, 
         ylim = ylim, 
         col = "red", 
         xlab = "time", 
         ylab = "temp",
         lwd = 2)
    
  }
  
  points(x, 
         y, 
         col = alpha("blue", 0.8),
         pch = 4)
}

x <- 0.4
y <- 0.719
sigN <- 0.1
sigF <- 1
l <- 0.3
x_star <- seq(-1, 1, length=100)

posterior <- posteriorGP(x, y, x_star, sigN, SquaredExpKernel, sigF, l)

plotGP(posterior$mean,
       posterior$var,
       x_star,
       x,
       y,
       title = "Posterior mean with 95% probability bands")

```

The probability bands is only close to the mean close to the given point. It is noit at the point because of assumed noise.

## (1.3)
Another point is added to see the effect on mean and their probability bands.
```{r}

x <- c(0.4, -0.6)
y <- c(0.719, -0.044)

posterior <- posteriorGP(x, y, x_star, sigN, SquaredExpKernel, sigF, l)

plotGP(posterior$mean,
       posterior$var, 
       x_star,
       x,
       y,
       title = "Posterior mean with 95% probability bands")

```

A similar response around the additional point as in **1.2**

## (1.4)
Even more points are added.
```{r}

x <- c(-1.0, -0.6, -0.2, 0.4, 0.8)
y <- c(0.768, -0.044, -0.940, 0.719, -0.664)

posterior <- posteriorGP(x, y, x_star, sigN, SquaredExpKernel, sigF, l)

plotGP(posterior$mean, 
       posterior$var, 
       x_star,
       x,
       y,
       title = "Posterior mean with 95% probability bands")

```

Points are now so close that the probability between them never reaches previous levels as they influence enough between eachother.

## (1.5)

```{r}

sigF <- 1
l <- 1

posterior <- posteriorGP(x, y, x_star, sigN, SquaredExpKernel, sigF, l)

plotGP(posterior$mean,
       posterior$var, 
       x_star,
       x,
       y,
       title = "Posterior mean with 95% probability bands")

```

A quite drastic change can be observed when having $\sigma_f$ and $l$ both equal 1 in the kernel function when comparing the graphs from **1.4** and **1.5**. Having a higher $l$ means that correlation between two adjacent points and gives a smoother function. So graph in **1.5** is more smooth.


## (2)
GP regression on temperature data using ``kernlab``

## (2.1)
A exponential kernel is created so that it can take in $\sigma_f$ and $l$
```{r}

KernelFunction <- function(sigmaF, l) {
  sqEx <- SquaredExpKernel <- function(x1, x2 = NULL){
    n1 <- length(x1)
    n2 <- length(x2)
    K <- matrix(NA,n1,n2)
    for (i in 1:n2){
      K[,i] <- sigmaF^2*exp(-0.5*( (x1-x2[i])/l)^2 )
    }
    return(K)
  }
  class(sqEx) <- "kernel"
  return(sqEx)
}

x <- c(1, 2, 3)
x_star <- c(2, 3, 4)

kernel <- KernelFunction(sigmaF = 1, l = 0.3)

kernel_test <- kernel(1, 2)
kernel_test

cov_matrix <- kernelMatrix(x, x_star)
cov_matrix

```

## (2.2)

```{r}

temps_csv <- read.csv("https://github.com/STIMALiU/AdvMLCourse/raw/master/GaussianProcess/Code/TempTullinge.csv", header=TRUE, sep=";")

by_ <- 5

time <- seq(1, 2190, by = by_)

temps <- temps_csv$temp[time]

fit <- lm(temps ~ time + I(time^2))

sigN <- sd(fit$residuals)
sigN
```
The estimate of the measurement noise ($\sigma_n$) is obtained by getting the standard deviation of the residuals when fitting a simple quadratic model to the data.
```{r}
sigF <- 20
l <- 0.2

GPfit <- gausspr(x = time, y = temps, kernel = KernelFunction, kpar = list(sigmaF = sigF, l = l), var = sigN^2)

GPmean <- predict(GPfit, time)

plotGP(mean = GPmean, interval = time, x = time, y = temps, title = "Posterior mean")

```

The posterior follows the data well. It does has a hard time coming close to extremes however.

## (2.3)

```{r}

kernel <- KernelFunction(sigmaF = sigmaF, l = l)

GPpred <- posteriorGP(scale(time), scale(temps), scale(time), sigN, SquaredExpKernel, sigmaF = sigF, l = l)

sdTemps <- sd(temps)
mnTemps <- mean(temps)

meanPred <- GPpred$mean * sdTemps + mnTemps

plotGP(mean = meanPred,
       GPpred$var, 
       interval = time, 
       x = time, 
       y = temps,
       title = "Time vs temp")

```

We can see that the bands incorporates more data points but fail to come close to the extremes (i.e at time ~750)

## (2.4)

```{r}

days <- rep(seq(1, 365, by = by_), times = 6)

GPdaily <- gausspr(days, 
                   temps, 
                   kernel = KernelFunction, 
                   kpar = list(sigmaF = sigF, l = l), 
                   var = sigN^2)

dailyMean <- predict(GPdaily, days)

plotGP(mean = GPmean, 
       var = GPpred$var, 
       interval = time, 
       x = time,
       y = temps, 
       title = "Time vs temp")

lines(x = time, y = dailyMean, col = "green", lwd = 2)

legend("bottomright", 
       legend = c("Data points", 
                  "Predicted mean (time)", 
                  "Predicted mean (day)", 
                  "95% probability bands"),
       col = c("blue", "red", "black", "green"),
       pch = c(4, NA, NA, NA),
       lwd = c(NA, 2, 2, 2))

```

Some points have improved when using the day set compared to the all-times set but they are mostly similar.

## (2.5)

```{r}

PeriodicKernel <- function(sigmaF, l1, l2, d){
  val <- function(x, x_star){
    dif <- abs(x - x_star)
    out <- sigmaF^2*exp(-((2*sin(pi*dif)) / d) / l1^2) * exp(-0.5 * dif^2 / l2^2)
    return(out)
  }
  class(val) <- "kernel"
  return(val)
}

l1 <- 1
l2 <- 10
d <- 365 / sd(time)

periodicKernel <- PeriodicKernel(sigF, l1, l2, d)
GPper <- gausspr(time, temps, kernel = periodicKernel, var = sigN^2)
perMean <- predict(GPper, time)
plotGP(mean = perMean,
       interval = time,
       x = time, 
       y = temps,
       title = "Time vs temp")

lines(x = time, y = GPmean, col = "black", lwd = 2)
lines(x = time, y = dailyMean, col = "green", lwd = 2)

legend("bottomright", 
       legend = c("Data points", 
                  "Periodic kernel",
                  "Predicted mean (time)", 
                  "Predicted mean (day)"),
       col = c("blue", "red", "black", "green"),
       pch = c(4, NA, NA, NA),
       lwd = c(NA, 2, 2, 2))



```
The periodic kernel is not as smooth as the other kernels but it does fit the data better. It captures more data further from the posterior mean.


## (3)

```{r}

library(AtmRay)

data <- read.csv("https://github.com/STIMALiU/AdvMLCourse/raw/master/GaussianProcess/Code/banknoteFraud.csv", header=FALSE, sep=",")
names(data) <- c("varWave","skewWave","kurtWave","entropyWave","fraud")
data[,5] <- as.factor(data[,5])
set.seed(111)
SelectTraining <- sample(1:dim(data)[1], size = 1000, replace = FALSE)

trainData <- data[SelectTraining, ]
testData <- data[-SelectTraining, ]

```

## (3.1)

```{r}

fraudModel <- gausspr(fraud ~ varWave + skewWave, data = trainData)

x1 <- seq(min(trainData$varWave), max(trainData$varWave), length=100)
x2 <- seq(min(trainData$skewWave), max(trainData$skewWave), length=100)

gridPoints <- meshgrid(x1, x2)
gridPoints <- cbind(c(gridPoints$x), c(gridPoints$y))
gridPoints <- data.frame(gridPoints)

names(gridPoints) <- c("varWave", "skewWave")

probPreds <- predict(fraudModel, gridPoints, type="probabilities")
trainPreds <- predict(fraudModel, trainData)

frauds <- which(trainData$fraud == 1)

contour(x1, 
        x2, 
        matrix(probPreds[, 2], 100, byrow = TRUE), 
        20, 
        xlab = "varWave", 
        ylab = "skewWave", 
        main = 'Fraud probs')

points(x = trainData$varWave[frauds], 
       y = trainData$skewWave[frauds], 
       col="blue")
points(x = trainData$varWave[-frauds], 
       y = trainData$skewWave[-frauds], 
       col="red")

confMatrix = table(trainPreds, trainData$fraud)
confMatrix
acc <- sum(diag(confMatrix)) / sum(confMatrix)
acc

```

The fraud points are red in the plot and the accuracy is 94.1%.

## (3.2)

```{r}

testPreds <- predict(fraudModel, newdata = testData)

confMatrix = table(testPreds, testData$fraud)
confMatrix
acc <- sum(diag(confMatrix)) / sum(confMatrix)
acc

```
The accuracy of the test data is high, but it naturally isn't as good as that of the training data.

## (3.3)

```{r}

fraudModel_2 <- gausspr(fraud ~., data = trainData)

testPreds_2 <- predict(fraudModel_2, newdata = testData)

confMatrix = table(testPreds_2, testData$fraud)
confMatrix
acc <- sum(diag(confMatrix)) / sum(confMatrix)
acc

```
Having more information leads to better accuracy since the new covariates only add additional information and do not remove any.
